\documentclass[fandol]{seuthesis-2022}

\title{基于领域自适应的}{工业过程故障诊断方法研究}
\author{陈鲲龙}
\studentnum{08022311}
\department{自动化}
\major{自动化}
\supervisor{王永健、童国道}
\period{2026年3月 \textasciitilde  2026年6月}

\begin{document}

\maketitle
\makedeclaration

\frontmatter
\begin{abstract}[zh]
工业过程故障诊断对于保障生产安全、降低非计划停机损失和提升过程运行可靠性具有重要意义。然而，在真实流程工业现场中，不同操作模式、负载水平和控制阶段往往对应不同的数据分布，使得仅依赖单一工况标注样本训练得到的模型难以直接迁移到新工况。面向这一问题，本文以 Tennessee Eastman Process Domain Adaptation 数据集为对象，研究单源到单目标的无监督领域自适应故障诊断，并围绕统一实验基准构建、基线方法比较和改进方法设计展开系统研究。

本文首先构建了面向 TEP 跨工况迁移诊断的统一实验流程。该流程以 manifest 驱动的数据解析方式管理 6 个运行模式、29 个类别和 5 折划分，在统一的一维全卷积网络骨干下，对 Source Only、CORAL、DAN、DANN、CDAN 和 DeepJDOT 等代表性方法进行可复现实验比较；同时统一了输入表示、域内标准化、优化器、模型选择指标、早停策略和结果产物格式，使不同方法之间的对比尽量聚焦于领域对齐机制本身。围绕四个关键单源场景，即 mode2$\rightarrow$mode5、mode3$\rightarrow$mode6、mode5$\rightarrow$mode2 和 mode6$\rightarrow$mode3，基线实验表明：相较于 Source Only 的平均目标域准确率 0.6706，CORAL、DAN、DANN、CDAN 和 DeepJDOT 的平均目标域准确率分别达到 0.7501、0.8118、0.8329、0.7852 和 0.7213，其中 DANN 在基线方法中平均表现最优，DAN 则展现出较好的跨场景稳定性。

在基线分析基础上，本文进一步提出可靠性门控的类条件时序自适应方法 RCTA。该方法以强基线对齐分支为底座，引入指数滑动平均教师网络、弱强双视图时序增强、多因素可靠性评分、按类门控伪标签采纳、源域与目标域双原型记忆以及原型吸引和一致性联合约束，用于缓解流程工业跨工况迁移中的伪标签噪声累积、全局对齐不足和类别混叠问题。实验结果显示，RCTA 在四个关键单源场景上均取得最高目标域准确率，平均目标域准确率达到 0.8590，相比 Source Only 平均提升 0.1884，相比逐场景最强基线平均进一步提升 0.0186，验证了该方法在流程工业无监督领域自适应故障诊断任务中的有效性。

总体而言，本文形成了一条从统一实验基准构建、典型基线比较到改进方法验证的完整研究链路。研究结果表明，在流程工业多类别时序迁移诊断任务中，仅依赖全局域对齐仍难以稳定解决类混叠与噪声伪监督问题，而将可靠性门控、教师--学生一致性与双原型结构结合起来，能够进一步提升跨工况迁移性能。

\keywords{故障诊断, 领域自适应, TEP, 无监督领域自适应}
\end{abstract}

\begin{abstract}[en]
\begingroup
\sloppy
Industrial fault diagnosis is important for production safety, reduction of unplanned downtime, and reliable process operation. In real process industries, however, data collected under different operating modes, load levels, or control stages usually follow different distributions. As a result, a model trained on labeled data from one operating condition often generalizes poorly to another condition. To address this problem, this thesis studies single-source to single-target unsupervised domain adaptation for fault diagnosis on the Tennessee Eastman Process Domain Adaptation dataset, with an emphasis on benchmark construction, unified comparison, and method improvement.

This work first establishes a reproducible experimental pipeline for cross-mode transfer diagnosis on TEP. A manifest-backed data interface is used to organize six operating modes, twenty-nine classes, and five folds. Under a shared one-dimensional fully convolutional backbone, six representative methods are compared: Source Only, CORAL, DAN, DANN, CDAN, and DeepJDOT. Input representation, domain-wise normalization, optimization settings, model-selection criteria, early stopping, and reporting artifacts are standardized so that the comparison focuses on the adaptation mechanisms themselves. On four key single-source scenarios, namely mode2$\rightarrow$mode5, mode3$\rightarrow$mode6, mode5$\rightarrow$mode2, and mode6$\rightarrow$mode3, the average target-domain accuracy reaches 0.7501, 0.8118, 0.8329, 0.7852, and 0.7213 for CORAL, DAN, DANN, CDAN, and DeepJDOT, respectively, compared with 0.6706 for the Source Only baseline.

Based on the baseline analysis, this thesis further proposes a Reliability-Gated Class-Conditional Temporal Adaptation method, abbreviated as RCTA. The method combines a strong base alignment branch with an exponential-moving-average teacher, conservative weak/strong temporal augmentations, reliability-aware class-wise pseudo-label gating, dual prototype memories for source and target domains, and prototype attraction plus consistency regularization. These mechanisms are designed to reduce noisy pseudo supervision, strengthen cross-domain intra-class aggregation, and alleviate class confusion caused by purely global alignment. Experimental results show that RCTA achieves the best target-domain accuracy in all four key scenarios, reaching an average of 0.8590. This corresponds to an average gain of 0.1884 over Source Only and 0.0186 over the strongest baseline on a per-scenario basis.

Overall, the thesis forms a coherent research line from benchmark construction to baseline comparison and improved-method validation. The results indicate that, for multi-class industrial time-series adaptation, global alignment alone is insufficient to stably resolve class confusion and noisy pseudo supervision, while the combination of reliability gating, teacher-student consistency, and dual-prototype structure can further improve cross-mode transfer performance.

\keywords{Fault Diagnosis, Domain Adaptation, TEP, Unsupervised Domain Adaptation}\par
\endgroup
\end{abstract}

\tableofcontents

\mainmatter
\input{chapters/01_intro}
\input{chapters/02_related_work}
\input{chapters/03_method}
\input{chapters/04_experiments}
\input{chapters/05_improved_method}
\input{chapters/06_conclusion}

\bibliographystyle{gbt7714-numerical}
\bibliography{bib/references}

\appendix
\input{chapters/appendix_a}

\begin{acknowledgement}
值此本科毕业论文完成之际，谨向在大学期间关心、指导和帮助过我的老师、学长、同学及家人致以诚挚的感谢。

首先，衷心感谢指导教师王永健教授，在毕业设计选题、研究思路梳理、实验推进和论文撰写过程中给予了我耐心细致的指导与帮助，使我得以顺利完成本论文的研究工作；同时，在个人发展与后续规划等方面给予我的支持与鼓励，也使我深受感激；严谨认真的治学态度和认真负责的工作作风，使我受益匪浅。

其次，感谢蒋徐颢、蔡闻骁、邹滨阳等学长和同学榜样在学习与成长过程中给予我的帮助与示范。他们在专业学习、科研视野和个人规划方面的经验分享，对我产生了积极而深远的影响。

感谢严震霆、周志铭、曲振宁等同学和朋友在本科阶段给予我的陪伴、支持与鼓励。在课题讨论、日常学习和生活交流中，他们给予了我许多真诚的帮助，使我能够更加从容地面对学习和毕业设计过程中的困难与挑战。

最后，衷心感谢我的父母和家人。感谢他们一直以来给予我的理解、包容和支持，使我能够安心完成本科阶段的学习任务与毕业论文工作。
\end{acknowledgement}

\end{document}
